---
title: "Mapping Discrete Traits onto Phylogeny"
author: "Callum Bucklow"
date: "`r Sys.Date()`"
output: html_document
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE) ##Remove messages for all chunks
knitr::opts_chunk$set(warning = FALSE) ##Remove warnings for all chunks
```

```{r Libraries}
# Load the necessary libraries
library(tidyverse) #data wrangling
library(tidyr) #data wrangling
library(dplyr)
library(viridis) #colour scale for heat maps
library(ggplot2) #for plotting when I don't want to use Base R
library(gridExtra) #for ggplot, to plot multiple plots together
library(fitdistrplus) #for fitting distributions to data
library(ape) #for phylogeny manipulation and all the rest
library(phytools) #for various things, including phylogenetic PCA, PGLM etc.
library(nlme) #for pgls
library(geiger) #for fitting trait evolution models to phylogenies
library(OUwie) #for fitting multiple regime OU models to trait evolution models
library(caper) #for pgls and will report pseudo-r2 for these models
```

# Defined Functions for Analysis

```{r Defining Dataframe Tree Pruning Function, echo=FALSE}

#Here a function is defined that automatically processes and prunes a phylogenetic tree and a corresponding dataframe (based on 'species_name') and reorders the dataframe to match the order of the tip.labels present on the phylogeny. The dataframe is subsequently converted into a trait matrix for input into various phytool (Revell) functions. It also stores qualitative data (e.g. mouthbrooder etc.) in a separate dataframe that can be accessed and appended to the PCA scores.

#note that it assumes that the tip.labels and species_names are formatted in the same way

prune_tree_and_dataframe <- function(input_tree, data_frame) {

  # Prune tree to species present in data frame
  pruned_tree <- drop.tip(input_tree, which(!input_tree$tip.label %in% data_frame$species_name))

  # Subset data frame to include only species present in the pruned tree
  pruned_data_frame <- data_frame[data_frame$species_name %in% pruned_tree$tip.label, ]

  # Reorder data frame based on the order of tip labels in the pruned tree
  pruned_data_frame <- pruned_data_frame[match(pruned_tree$tip.label, pruned_data_frame$species_name), ]

  # Store columns 1:9 from the pruned data frame
  stored_data <- pruned_data_frame[, 1:9]

  # Remove columns 1:9 and 14 from the pruned data frame
  pruned_data_frame <- pruned_data_frame[, -c(1:9, 14)]

  # Convert pruned data frame into a matrix
  pruned_matrix <- as.matrix(pruned_data_frame)

  # Use the first column of stored_data as row names for the matrix
  rownames(pruned_matrix) <- stored_data[, 1]

  return(list(tree = pruned_tree, matrix_data = pruned_matrix, stored_data = stored_data))
}

```

# Data Import

```{r Import All Species RDS Object}
all_species <- readRDS("Data/all_species.rds")
```

# Prepare All Species Data

```{r Preparing All Species Data}
all_species_stored_data <- all_species$stored_data
all_species_matrix_data <- all_species$matrix_data
all_species_tree <- all_species$tree

rownames(all_species_stored_data) <- all_species_stored_data$species_name

# Convert matrix to a dataframe
all_species_matrix_data_df <- as.data.frame(all_species_matrix_data)
```

# Mapping Piscivory onto Whole Phylogeny (Discrete)

```{r Prepare All Species Data to Remove Data Deficients for Diet}
#Filter data for piscviory/diet preference:
all_species_stored_data_piscivore <- 
  all_species_stored_data %>%
  filter(piscivore != "data_deficient")

#save the qualitative data
write.csv(all_species_stored_data_piscivore, 'Data/all_species_stored_data_piscivore.csv')

# Filter rows based on row names not present in another matrix
all_species_matrix_data_piscivore <- all_species_matrix_data_df %>%
   filter((row.names(all_species_matrix_data_df) %in% row.names(all_species_stored_data_piscivore)))

#save the quantitative data
write.csv(all_species_matrix_data_piscivore, 'Data/all_species_matrix_data_piscivore.csv')

#prune the tree:
all_species_tree_piscivore_pruned <- drop.tip(all_species_tree, which(!all_species_tree$tip.label %in% rownames(all_species_matrix_data_piscivore)))

#save pruned tree:
write.tree(all_species_tree_piscivore_pruned, 'Data/all_species_tree_piscivore_pruned.txt')

#save edge lengths for later:
all_species_tree_piscivore_pruned_edges <- all_species_tree_piscivore_pruned$edge.length

#create trait variable for fitting discrete models:
all_species_piscivore <- all_species_stored_data_piscivore$piscivore
names(all_species_piscivore) <- row.names(all_species_stored_data_piscivore)


all_species_tree_piscivore_pruned
```

```{r Fit Piscivore Discrete Character Trait Evolution Models for All Species, echo=FALSE, eval=FALSE}
# Fit models of discrete character evolution. Note that since the trait we are interested in is binary, we do not need to also fit a symmetrical (SYM) model to the data as it reduces to an equal rates model for k = 2. 
#fit_ER_all_species_piscivore <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_piscivore, model = 'ER')


#fit_ARD_all_species_piscivore <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_piscivore, model = 'ARD')

#compare the fit of the three models:
#aicw(c("ER" = fit_ER_all_species_piscivore$opt$aicc,
       #"ARD" = fit_ARD_all_species_piscivore$opt$aicc))
#ARD model fits the data best
```

```{r Fit ARD Model for Piscivore Behaviour to All Species Tree}
#Rescale all branches of the tree to 1 to avoid issues with
#all_species_tree_piscivore_pruned$edge.length <- all_species_tree_piscivore_pruned$edge.length/all_species_tree_piscivore_pruned$edge.length

#Fit ARD model for piscivorous behaviour for all species onto the tree:
#all_species_piscivore_simmap <- make.simmap(tree = all_species_tree_piscivore_pruned, x = all_species_piscivore, model = 'ARD', nsim = 10000)

#save the multiphylo simmap object so analysis doesn't need to be redone:
#saveRDS(all_species_piscivore_simmap, 'Data/all_species_piscivore_simmap.rds')

#store all the trees simulated in the simmap creation, this will make it easier to plot the uncertainty at nodes etc.
#posterior_all_species_piscivore_simmap <- 
  #describe.simmap(all_species_piscivore_simmap, plot = FALSE)

# Save simmap object as RDS - this is important if the simmap requires output for further analysis
#saveRDS(posterior_all_species_piscivore_simmap, 'Data/posterior_all_species_piscivore_simmap.rds')
```

```{r Plot Simmap object for Piscivory for All Species}

#import the simmap object:
#posterior_all_species_piscivore_simmap <- readRDS("Data/posterior_all_species_piscivore_simmap.rds")

#replace branch lengths to those of the original tree for each tree:
#for (i in 1:length(posterior_all_species_piscivore_simmap$tree)) {
  #posterior_all_species_piscivore_simmap$tree[[i]]$edge.length <- all_species_tree_piscivore_pruned_edges
#}

#defining tip colours vector
tip_colours_vector <- as.data.frame(all_species_piscivore)
tip_colours_vector$species_name <- rownames(tip_colours_vector)
rownames(tip_colours_vector) <- NULL

#extract tip labels from the pruned tree
tip_labels <- all_species_tree_piscivore_pruned$tip.label

#create a named vector of colors matching the tip labels
tip_colours_vector <- ifelse(
  tip_labels %in% tip_colours_vector$species_name[tip_colours_vector$all_species_piscivore == "non-piscivore"],
  "darkgreen",
  "purple"
)

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_all_species_piscivore_simmap.pdf", width = 8.4, height = 11)

#plot reconstruction with posterior probabilities at each node
#define the piscivory colours for plotting onto the phylogeny
piscivore_colours <- c("non-piscivore" = 'darkgreen',  "piscivore" = 'purple')

plot(all_species_tree_piscivore_pruned,
     type="fan",
     cex=0.25,
     label.offset = 0.5,
     tip.color = tip_colours_vector)
nodelabels(pie=posterior_all_species_piscivore_simmap$ace,piecol=piscivore_colours,
cex=0.25)
# Add the legend
add.simmap.legend(colors=piscivore_colours, 
                  prompt=FALSE, 
                  fsize=0.75,     
                  size=10, 
                  x = 1,       
                  y = 20)        
add.scale.bar()

dev.off()
```







```{r Prune Simmap object for Piscivory to Lake Malawi}

all_species_piscivore_simmap <- readRDS('Data/all_species_piscivore_simmap.rds')

#replace branch lengths to those of the original tree for each tree:
for (i in seq_along(all_species_piscivore_simmap)) {
  all_species_piscivore_simmap[[i]]$edge.length <- all_species_tree_piscivore_pruned_edges
}

lake_malawi_trees <- lapply(all_species_piscivore_simmap,extract.clade.simmap,node=433)
class(lake_malawi_trees)<-"multiPhylo"

lake_malawi_described_simmap <- describe.simmap(lake_malawi_trees, plot = FALSE)

lake_malawi_tree_pruned <- extract.clade(all_species_tree_piscivore_pruned, node=433)
```

```{r Plot Simmap object for Piscivory for Lake Malawi}

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_lake_malawi_piscivore_simmap.pdf", width = 8.4, height = 11)

# Plot reconstruction with posterior probabilities at each node
#define the piscivory colours for plotting onto the phylogeny
piscivore_colours <- c("non-piscivore" = 'darkgreen',  "piscivore" = 'purple')

plot(lake_malawi_tree_pruned,
     type="fan",
     cex=0.33,
     label.offset = 0.05)
nodelabels(pie=lake_malawi_described_simmap$ace,piecol=piscivore_colours,
cex=0.40)
# Add the legend
add.simmap.legend(colors=piscivore_colours, 
                  prompt=FALSE, 
                  fsize=0.75,     
                  size=5, 
                  x = 0,       
                  y = 0.50)        
#add.scale.bar()

dev.off()
```

```{r Calculate Piscivore Discrete Trait Evolution Density Map}
#since the trait is binary, we can probably plot it as a density map:

#import the simmap object:
#all_species_piscivore_simmap <- readRDS("Data/all_species_piscivore_simmap.rds")

#create the density map object:
#density_map_all_species_piscivore <- densityMap(all_species_piscivore_simmap,
                                                #plot = FALSE)

#save as RDS object to prevent need to redo analysis:
#saveRDS(density_map_all_species_piscivore, "Data/density_map_all_species_piscivore.rds")

```

```{r Plot Piscivore Discrete Trait Evolution as a Density Map}

#import the density map RDS object:
#density_map_all_species_piscivore <- readRDS("Data/density_map_all_species_piscivore.rds")

#set the mapping variables (i.e. variable colour)
#density_map_all_species_piscivore <- setMap(density_map_all_species_piscivore, 
                                              #piscivore_colours)

#replace branch lengths from 1 back to original lengths:
#density_map_all_species_piscivore$tree$edge.length <- all_species_tree_piscivore_pruned_edges

#plot the density map:
#pdf("Figures/Mapping Discrete Traits onto Phylogeny/density_map_all_species_piscivore.pdf", width = 8.4, height = 11)

#plot empty density map:
#plot(density_map_all_species_piscivore,
     #type="fan",
     #lwd=c(1, 5),
     #tip.label = NA,
     #ftype = "i",
     #fsize = 0.25)

#dev.off()
```

# Mapping Water System onto Whole Phylogeny (Discrete)

```{r Prepare All Species Data for Water System Discrete Modelling}

#we have missing data for piscivory, to make sure trees are comparable we should fit all data to minimal piscivore tree:

#create trait variable for fitting discrete models:
all_species_water_system <- all_species_stored_data_piscivore$water_system
names(all_species_water_system) <- all_species_stored_data_piscivore$species_name

#save edge lengths for later:
all_species_tree_water_system_pruned_edges <- all_species_tree_piscivore_pruned$edge.length
```

```{r Fit Water System Discrete Character Trait Evolution Models for All Species, echo=FALSE, eval=FALSE}
# Fit models of discrete character evolution. Note that since the trait we are interested in is NOT binary, we should also test the fit of a symmetrical model (sym) in which forward and reverse transitions share the same parameter
# fit_ER_all_species_water_system <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_water_system, model = 'ER')
# 
# fit_ARD_all_species_water_system <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_water_system, model = 'ARD')
# 
# fit_SYM_all_species_water_system <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_water_system, model = 'SYM')
# 
# #compare the fit of the three models:
# aicw(c("ER" = fit_ER_all_species_water_system$opt$aicc,
#        "ARD" = fit_ARD_all_species_water_system$opt$aicc,
#        "SYM" = fit_SYM_all_species_water_system$opt$aicc))
# #ARD model fits the data best
```


```{r Fit ARD Model for Water System to All Species Tree, echo=FALSE, eval=FALSE}
#Rescale all branches of the tree to 1 to avoid issues with
# all_species_tree_piscivore_pruned$edge.length <- all_species_tree_piscivore_pruned$edge.length/all_species_tree_piscivore_pruned$edge.length
# 
# #Fit ARD model for water system occupancy for all species onto the tree:
# all_species_water_system_simmap <- make.simmap(tree = all_species_tree_piscivore_pruned, x = all_species_water_system, model = 'ARD', nsim = 10000)
# 
# #save the multiphylo simmap object so analysis doesn't need to be redone:
# saveRDS(all_species_water_system_simmap, 'Data/all_species_water_system_simmap.rds')
# 
# #store all the trees simulated in the simmap creation, this will make it easier to plot the uncertainty at nodes etc.
# posterior_all_species_water_system_simmap <- 
#   describe.simmap(all_species_water_system_simmap, plot = FALSE)
# 
# # Save simmap object as RDS - this is important if the simmap requires output for further analysis
# saveRDS(posterior_all_species_water_system_simmap, 'Data/posterior_all_species_water_system_simmap.rds')
```

```{r Prepare Simmap object for Water System for All Species}

#import the simmap object:
posterior_all_species_water_system_simmap <- readRDS("Data/posterior_all_species_water_system_simmap.rds")

#replace branch lengths to those of the original tree for each tree - probably not necessary.
for (i in 1:length(posterior_all_species_water_system_simmap$tree)) {
  posterior_all_species_water_system_simmap$tree[[i]]$edge.length <- all_species_tree_water_system_pruned_edges
}
```

```{r Plot Simmap object for Water System for All Species}

#create new tree object and switch edge lengths back to original tree (since we updated them to one)
all_species_tree_piscivore_pruned$edge.length <- all_species_tree_water_system_pruned_edges

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_all_species_water_system_simmap.pdf", width = 8.4, height = 11)

# Plot reconstruction with posterior probabilities at each node
#colours taken from Wong colour blind palette
water_system_colours <- c("Lake Malawi" = "#D55E00",
                          "Lake Tanganyika" = "#0072B2",
                          "Lake Victoria" = "#43AA8B",
                          "Riverine" = "#CC79A7")

#defining tip colours vector
tip_colours_vector <- as.data.frame(all_species_water_system)
tip_colours_vector$species_name <- rownames(tip_colours_vector)
rownames(tip_colours_vector) <- NULL

#create a vector of species names (you would replace this with your actual list of species)
species_names <- tip_colours_vector$species_name

#create a vector of water systems for each species (replace with your actual data)
species_water_system <- tip_colours_vector$all_species_water_system
species_water_system <- as.character(species_water_system)

#convert this to a named vector for easy lookup
species_to_water_system <- setNames(species_water_system, species_names)

#map the water system colours based on the species
tip_colours_vector <- water_system_colours[species_to_water_system[species_names]]


plot(all_species_tree_piscivore_pruned,
     type="fan",
     cex=0.25,
     label.offset = 0.5,
     tip.color = tip_colours_vector)
nodelabels(pie=posterior_all_species_water_system_simmap$ace,piecol=water_system_colours,
cex=0.25)
add.simmap.legend(colors=water_system_colours,                   
                  prompt=FALSE, 
                  fsize=0.75,  
                  size=10, 
                  x = 1,      
                  y = 20)
add.scale.bar()

dev.off()
```

# Mapping Water System onto Whole Phylogeny as Riverine and Lacustrine

```{r Changing Water Systems to Lacustrine and Riverine}

#create a copy of all_species_water_system
all_species_riverine_lacustrine <- all_species_water_system

#convert to character to avoid factor level issues
all_species_riverine_lacustrine <- as.character(all_species_riverine_lacustrine)

#replace specific lake names with 'Lacustrine'
all_species_riverine_lacustrine[all_species_riverine_lacustrine %in% c("Lake Malawi", "Lake Tanganyika", "Lake Victoria")] <- "Lacustrine"

#redefine the rownames
names(all_species_riverine_lacustrine) <- names(all_species_water_system)

#rescale all branches of the tree to 1 to avoid issues with simmap estimation
all_species_tree_riverine_lacustrine <- all_species_tree_piscivore_pruned
all_species_tree_riverine_lacustrine$edge.length <- all_species_tree_riverine_lacustrine$edge.length/all_species_tree_riverine_lacustrine$edge.length
```

```{r Fit ARD Model for Riverine or Lacustrine to All Species Tree, echo=FALSE, eval=FALSE}
#Fit ARD model for riverine or lacustrine for all species onto the tree:
# all_species_riverine_lacustrine_simmap <- make.simmap(tree = all_species_tree_piscivore_pruned, x = all_species_riverine_lacustrine, model = 'ARD', nsim = 10000)  
# 
# #save the multiphylo simmap object so analysis doesn't need to be redone:
# saveRDS(all_species_riverine_lacustrine_simmap, 'Data/all_species_riverine_lacustrine_simmap.rds')
# 
# #store all the trees simulated in the simmap creation, this will make it easier to plot the uncertainty at nodes etc.
# posterior_all_species_riverine_lacustrine_simmap <- 
#   describe.simmap(all_species_riverine_lacustrine_simmap, plot = FALSE)
# 
# #Save simmap object as RDS - this is important if the simmap requires output for further analysis
# saveRDS(posterior_all_species_riverine_lacustrine_simmap, 'Data/posterior_all_species_riverine_lacustrine_simmap.rds')
```

```{r Prepare Simmap object for Lacustrine or Riverine for All Species}

#import the simmap object:
posterior_all_species_riverine_lacustrine_simmap <- readRDS("Data/posterior_all_species_riverine_lacustrine_simmap.rds")

#replace branch lengths to those of the original tree for each tree - probably not necessary.
for (i in 1:length(posterior_all_species_riverine_lacustrine_simmap$tree)) {
  posterior_all_species_riverine_lacustrine_simmap$tree[[i]]$edge.length <- all_species_tree_water_system_pruned_edges
}
```

```{r Plot Simmap object for Riverine or Lacustrine for All Species}

#create new tree object and switch edge lengths back to original tree (since we updated them to one)
all_species_tree_piscivore_pruned$edge.length <- all_species_tree_water_system_pruned_edges

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_all_species_riverine_lacustrine_simmap.pdf", width = 8.4, height = 11)

# Plot reconstruction with posterior probabilities at each node
#colours taken from Wong colour blind palette
water_system_colours <- c("Lacustrine" = "#56B4E9",
                          "Riverine" = "#CC79A7")

#defining tip colours vector
tip_colours_vector <- as.data.frame(all_species_riverine_lacustrine)
tip_colours_vector$species_name <- rownames(tip_colours_vector)
rownames(tip_colours_vector) <- NULL

#extract tip labels from the pruned tree
tip_labels <- all_species_tree_piscivore_pruned$tip.label

#create a named vector of colors matching the tip labels
tip_colours_vector <- ifelse(
  tip_labels %in% tip_colours_vector$species_name[tip_colours_vector$all_species_riverine_lacustrine == "Lacustrine"],
  "#56B4E9",
  "#CC79A7"
)

plot(all_species_tree_piscivore_pruned,
     type="fan",
     cex=0.25,
     label.offset = 0.5,
     tip.color = tip_colours_vector)
nodelabels(pie=posterior_all_species_riverine_lacustrine_simmap$ace,piecol=water_system_colours,
cex=0.25)
add.simmap.legend(colors=water_system_colours,                   
                  prompt=FALSE, 
                  fsize=0.75,  
                  size=10, 
                  x = -0.50,      
                  y = 10)  
add.scale.bar(x = 0,
              y = 2)

dev.off()
```

# Mapping Depth Preference onto Whole Phylogeny (Discrete)

```{r Prepare All Species Data with their Depth Preference}
#we have missing data for piscivory, to make sure trees are comparable we should fit all data to minimal piscivore tree:

#create trait variable for fitting discrete models:
all_species_depth <- all_species_stored_data_piscivore$depth
names(all_species_depth) <- all_species_stored_data_piscivore$species_name

#save edge lengths for later:
all_species_tree_depth_pruned_edges <- all_species_tree_piscivore_pruned$edge.length

```

```{r Fit Depth Preference Discrete Character Trait Evolution Models for All Species}
# Fit models of discrete character evolution. Note that since the trait we are interested in is NOT binary, we should also test the fit of a symmetrical model (sym) in which forward and reverse transitions share the same parameter
# fit_ER_all_species_depth <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_depth, model = 'ER')
# 
# fit_ARD_all_species_depth <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_depth, model = 'ARD')
# 
# fit_SYM_all_species_depth <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_depth, model = 'SYM')
# 
# #compare the fit of the three models:
# aicw(c("ER" = fit_ER_all_species_depth$opt$aicc,
#        "ARD" = fit_ARD_all_species_depth$opt$aicc,
#        "SYM" = fit_SYM_all_species_depth$opt$aicc))
#ARD model fits the data best
```

```{r Fit ARD Model for Depth Preference to All Species Tree}

#Rescale all branches of the tree to 1 to avoid issues with
# all_species_tree_piscivore_pruned$edge.length <- all_species_tree_piscivore_pruned$edge.length/all_species_tree_piscivore_pruned$edge.length
# 
# #Fit ARD model for piscivorous behaviour for all species onto the tree:
# all_species_depth_simmap <- make.simmap(tree = all_species_tree_piscivore_pruned, 
#                                            x = all_species_depth, 
#                                            model = 'ARD', nsim = 10000)
# 
# #save simmap object to prevent needing to redo the analysis:
# saveRDS(all_species_depth_simmap, 'Data/all_species_depth_simmap.rds')
# 
# #store all the trees simulated in the simmap creation, this will make it easier to plot the uncertainty at nodes etc.
# posterior_all_species_depth_simmap <- 
#   describe.simmap(all_species_depth_simmap, plot = FALSE)
# 
# # Save simmap object as RDS - this is important if the simmap requires output for further analysis
# saveRDS(posterior_all_species_depth_simmap, 'Data/posterior_all_species_depth_simmap.rds')
```

```{r Plot Simmap object for Depth Preference for All Species}

#import the simmap with all 10000 trees:
posterior_all_species_depth_simmap <-
  readRDS('Data/posterior_all_species_depth_simmap.rds')

#define colours for each depth preference
depth_colours <- c("bathydemersal" = "#D81B60",
                                "benthopelagic" = "#1E88E5",
                                "demersal" = "#FFC107", 
                                "pelagic" = "#004D40") 

#defining tip colours vector
tip_colours_vector <- as.data.frame(all_species_depth)
tip_colours_vector$species_name <- rownames(tip_colours_vector)
rownames(tip_colours_vector) <- NULL

# Match tip labels directly to their corresponding colors
species_to_type <- setNames(tip_colours_vector$all_species_depth, tip_colours_vector$species_name)
tip_colours_vector <- depth_colours[species_to_type[all_species_tree_piscivore_pruned$tip.label]]

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_all_species_depth_simmap.pdf", width = 8.4, height = 11)

#replace branch lengths back to original (remember we scaled to 1 for model fit)
all_species_tree_piscivore_pruned$edge.length <- all_species_tree_depth_pruned_edges

# Plot reconstruction with posterior probabilities at each node
plot(all_species_tree_piscivore_pruned,
     type="fan",
     cex=0.25,
     label.offset = 0.5,
     tip.color = tip_colours_vector)
nodelabels(pie=posterior_all_species_depth_simmap$ace,piecol=depth_colours,
cex=0.25)

# Add the legend
add.simmap.legend(colors=depth_colours, 
                  prompt=FALSE, 
                  fsize=0.75,     
                  size=10, 
                  x = 1,       
                  y = 20)        
add.scale.bar()

dev.off()
```

```{r Prune Simmap object for Depth Preference to Lake Malawi}

#all_species_depth_simmap <- readRDS('Data/all_species_depth_simmap.rds')

#replace branch lengths to those of the original tree for each tree:
#for (i in seq_along(all_species_depth_simmap)) {
  #all_species_depth_simmap[[i]]$edge.length <- all_species_tree_depth_pruned_edges
#}

#lake_malawi_trees <- lapply(all_species_depth_simmap,extract.clade.simmap,node=447)
#class(lake_malawi_trees)<-"multiPhylo"

#lake_malawi_described_simmap <- describe.simmap(lake_malawi_trees, plot = FALSE)

#lake_malawi_tree_pruned <- extract.clade(all_species_tree, node=447)
```

```{r Plot Simmap object for Depth Preference for Lake Malawi}

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_lake_malawi_depth_simmap.pdf", width = 8.4, height = 11)

# Plot reconstruction with posterior probabilities at each node
#define the piscivory colours for plotting onto the phylogeny
#colours taken from Wong colour blind palette
depth_colours <- c("bathydemersal" = "blue",
                                "benthopelagic" = "purple", 
                                "demersal" = "red", 
                                "pelagic" = "green")
 
plot(lake_malawi_tree_pruned,
     type="fan",
     cex=0.33,
     label.offset = 0.05)
nodelabels(pie=lake_malawi_described_simmap$ace,piecol=depth_colours,
cex=0.40)
# Add the legend
add.simmap.legend(colors=depth_colours, 
                  prompt=FALSE, 
                  fsize=0.75,     
                  size=5, 
                  x = 0,       
                  y = 0.50)        
#add.scale.bar()

dev.off()
```

# Fitting Discrete Model of Trait Evolution for Haplochromine, LT-endemic and (Basal) Riverine

```{r Prepare All Species Data with their Tribe}
#we have missing data for piscivory, to make sure trees are comparable we should fit all data to minimal piscivore tree:

#create trait variable for fitting discrete models:
all_species_tribe <- all_species_stored_data_piscivore$tribe
names(all_species_tribe) <- all_species_stored_data_piscivore$species_name

#collapse the tribes into basal riverine, haplochromini or lake_tanganyika:
#define the haplochromini tribes/subtribes
haplochromini <- c("Basal Haplochromini", "Haplochromini", "Basal Modern Haplochromini", "Tropheini", "Serranochromini")

#define tribes endemic to LT but that are not haplochromines
lake_tanganyika <- c("Benthochromini", "Perissodini", "Cyprichromini",  "Ectodini", "Limnochromini",  "Cyphotilapiini", "Trematocarini", "Eretmodini",  "Lamprologini", "Bathybatini", "Boulengerochromini")

#define tribes endemic to riverines but not haplochromines
basal_riverine <- c("Chromidotilapiini", "Hemichromini", "Pelmatochromini", "Coptodonini", "Gobiocichlini", "Steatocranini",  "Etiini", "Tilapiini",  "Pelmatolapiini", "Oreochromini", "Tylochromini", "Heterochromini")

#create mapping vector
tribe_category <- case_when(
  all_species_tribe %in% haplochromini ~ "haplochromini",
  all_species_tribe %in% lake_tanganyika ~ "lake_tanganyika",
  all_species_tribe %in% basal_riverine ~ "basal_riverine",
  TRUE ~ "unknown" # Fallback in case a tribe is not categorized
)

#redine all_species_tribe
all_species_tribe <- tribe_category
names(all_species_tribe) <- all_species_stored_data_piscivore$species_name

#save edge lengths for later:
all_species_tree_tribe_pruned_edges <- all_species_tree_piscivore_pruned$edge.length
```

```{r Fit Tribal Discrete Character Trait Evolution Models for All Species}
##fit models of discrete character evolution. Note that since the trait we are interested in is NOT binary, we should also test the fit of a symmetrical model (sym) in which forward and reverse transitions share the same parameter
 #fit_ER_all_species_tribe<- fitDiscrete(all_species_tree_piscivore_pruned, all_species_tribe, model = 'ER')
# 
 #fit_ARD_all_species_tribe <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_tribe, model = 'ARD')
# 
 #fit_SYM_all_species_tribe <- fitDiscrete(all_species_tree_piscivore_pruned, all_species_tribe, model = 'SYM')
# 
# #compare the fit of the three models:
 #aicw(c("ER" = fit_ER_all_species_tribe$opt$aicc,
        #"ARD" = fit_ARD_all_species_tribe$opt$aicc,
        #"SYM" = fit_SYM_all_species_tribe$opt$aicc))
# #ER model fits the data best
```
```{r Fit ER Model for Tribe to All Species Tree}

#Rescale all branches of the tree to 1 to avoid issues with
#all_species_tree_piscivore_pruned$edge.length <- all_species_tree_piscivore_pruned$edge.length/all_species_tree_piscivore_pruned$edge.length

#Fit ARD model for tribe for all species onto the tree:
#set.seed(1931)
#all_species_tribe_simmap <- make.simmap(tree = all_species_tree_piscivore_pruned, 
                                           #x = all_species_tribe, 
                                           #model = 'ER', nsim = 10000)

#save simmap object to prevent needing to redo the analysis:
#saveRDS(all_species_tribe_simmap, 'Data/all_species_tribe_simmap.rds')

#store all the trees simulated in the simmap creation, this will make it easier to plot the uncertainty at nodes etc.
#posterior_all_species_tribe_simmap <- 
  #describe.simmap(all_species_tribe_simmap, plot = FALSE)

#Save simmap object as RDS - this is important if the simmap requires output for further analysis
#saveRDS(posterior_all_species_tribe_simmap, 'Data/posterior_all_species_tribe_simmap.rds')
```

```{r Plot Simmap object for Brooding Behaviour for All Species}

#import the simmap with all 10000 trees:
#posterior_all_species_tribe_simmap <- 
  #readRDS('Data/posterior_all_species_tribe_simmap.rds')

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_all_species_tribe_simmap.pdf", width = 8.4, height = 11)

#replace branch lengths back to original (remember we scaled to 1 for model fit)
all_species_tree_piscivore_pruned$edge.length <- all_species_tree_tribe_pruned_edges

# Plot reconstruction with posterior probabilities at each node
#colours taken from Wong colour blind palette
tribe_colours <- c("Basal Riverine" = "#CC79A7",
                   "Haplochromini" = "#D55E00",
                   "Non-Haplochromines (LT)" = "#0072B2")

#redefine factors:
names(posterior_all_species_tribe_simmap$ace) <- c("Basal Riverine", "Haplochromini", "Non-Haplochromines (LT)")

plot(all_species_tree_piscivore_pruned,
     type="fan",
     cex=0.25,
     label.offset = 0.5)
nodelabels(pie=posterior_all_species_tribe_simmap$ace,piecol=tribe_colours,
cex=0.25)
# Add the legend
add.simmap.legend(colors=tribe_colours, 
                  prompt=FALSE, 
                  fsize=0.75,  
                  size=10, 
                  x = 1,      
                  y = 20)       
add.scale.bar()

dev.off()
```









# Fitting Discrete Model of Trait Evolution for Brooding Behaviour

```{r Prepare All Species Data to Remove Species with Unknown Brooding Behaviour}

#filter out species for which we do not have data on brooding behaviour
all_species_stored_data_brooding <- 
  all_species_stored_data %>%
  filter(mouthbrooder != "data_deficient")

# Filter rows based on row names not present in another matrix to remove species with 'data_deficient' entries
all_species_matrix_data_brooding <- all_species_matrix_data_df %>%
   filter((row.names(all_species_matrix_data_df) %in% row.names(all_species_stored_data_brooding)))

#create trait variable for fitting discrete models:
all_species_brooding <- all_species_stored_data_brooding$mouthbrooder
names(all_species_brooding) <- row.names(all_species_stored_data_brooding)

#prune the tree:
all_species_tree_brooding_pruned <- drop.tip(all_species_tree, which(!all_species_tree$tip.label %in% rownames(all_species_matrix_data_brooding)))

#save edge lengths for later:
all_species_tree_brooding_pruned_edges <- all_species_tree_brooding_pruned$edge.length

#create trait variable for fitting discrete models:
all_species_brooding <- all_species_stored_data_brooding$mouthbrooder
names(all_species_brooding) <- row.names(all_species_stored_data_brooding)
```

```{r Fit Brooding Behaviour Discrete Character Trait Evolution Models for All Species}
# #fit models of discrete character evolution. Note that since the trait we are interested in is NOT binary, we should also test the fit of a symmetrical model (sym) in which forward and reverse transitions share the same parameter
# fit_ER_all_species_brooding <- fitDiscrete(all_species_tree_brooding_pruned, all_species_brooding, model = 'ER')
# 
# fit_ARD_all_species_brooding <- fitDiscrete(all_species_tree_brooding_pruned, all_species_brooding, model = 'ARD')
# 
# fit_SYM_all_species_brooding <- fitDiscrete(all_species_tree_brooding_pruned, all_species_brooding, model = 'SYM')
# 
# #compare the fit of the three models:
# aicw(c("ER" = fit_ER_all_species_brooding$opt$aicc,
#        "ARD" = fit_ARD_all_species_brooding$opt$aicc,
#        "SYM" = fit_SYM_all_species_brooding$opt$aicc))
# #ARD model fits the data best
```

```{r Fit ARD Model for Brooding Behaviour to All Species Tree}

#Rescale all branches of the tree to 1 to avoid issues with
all_species_tree_brooding_pruned$edge.length <- all_species_tree_brooding_pruned$edge.length/all_species_tree_brooding_pruned $edge.length

#Fit ARD model for brooding behaviour for all species onto the tree:
set.seed(1931)
all_species_brooding_simmap <- make.simmap(tree = all_species_tree_brooding_pruned, 
                                           x = all_species_brooding, 
                                           model = 'ARD', nsim = 10000)

#save simmap object to prevent needing to redo the analysis:
saveRDS(all_species_brooding_simmap, 'Data/all_species_brooding_simmap.rds')

#store all the trees simulated in the simmap creation, this will make it easier to plot the uncertainty at nodes etc.
posterior_all_species_brooding_simmap <- 
  describe.simmap(all_species_brooding_simmap, plot = FALSE)

#Save simmap object as RDS - this is important if the simmap requires output for further analysis
saveRDS(posterior_all_species_brooding_simmap, 'Data/posterior_all_species_brooding_simmap.rds')
```

```{r Plot Simmap object for Brooding Behaviour for All Species}

#import the simmap with all 10000 trees:
posterior_all_species_brooding_simmap <- 
  readRDS('Data/posterior_all_species_brooding_simmap.rds')

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_all_species_brooding_simmap.pdf", width = 8.4, height = 11)

#replace branch lengths back to original (remember we scaled to 1 for model fit)
all_species_tree_brooding_pruned$edge.length <- all_species_tree_brooding_pruned_edges

# Plot reconstruction with posterior probabilities at each node
#colours taken from Wong colour blind palette
brooding_behaviour_colours <- c("biparental_mouthbrooder" = "purple",
                                "maternal_mouthbrooder" = "red", 
                                "paternal_mouthbrooder" = "blue", 
                                "substrate_brooder" = "green"
                                )

 
plot(all_species_tree_brooding_pruned,
     type="fan",
     cex=0.25,
     label.offset = 0.5)
nodelabels(pie=posterior_all_species_brooding_simmap$ace,piecol=brooding_behaviour_colours,
cex=0.25)
# Add the legend
add.simmap.legend(colors=brooding_behaviour_colours, 
                  prompt=FALSE, 
                  fsize=0.75,  
                  size=10, 
                  x = 1,      
                  y = 20)       
add.scale.bar()

dev.off()
```

#Fitting Models of Discrete Trait Evolution Mouthbrooding versus Substrate Brooding (Binary Brooding Behaviour)

```{r Respecify Brooding Behaviour to Binary Trait (Substrate versus Mouthbrooding}

#specify new vector where we reduce three types of mouthbrooding to just mouthbrooding, so we have vector of mouthbrooding versus substrate brooding
all_species_brooding_mb_sb <- gsub("maternal_mouthbrooder|paternal_mouthbrooder|biparental_mouthbrooder", "mouthbrooder", all_species_brooding)

```

```{r Discrete Trait Evolution Model Fits for Binary Brooding Behaviour for All Species}

# Fit models of discrete character evolution. Note that since the trait we are interested is NOT binary, we do not need to test for sym model
fit_ER_all_species_brooding_mb_sb <- fitDiscrete(all_species_tree_brooding_pruned, all_species_brooding_mb_sb, model = 'ER')

fit_ARD_all_species_brooding_mb_sb <- fitDiscrete(all_species_tree_brooding_pruned, all_species_brooding_mb_sb, model = 'ARD')

#compare the fit of the three models:
aicw(c("ER" = fit_ER_all_species_brooding_mb_sb$opt$aicc,
       "ARD" = fit_ARD_all_species_brooding_mb_sb$opt$aicc))
#ARD model fits the data best
```

```{r Fit ARD Model for Binary Brooding Behaviour to All Species Tree}

#Rescale all branches of the tree to 1 to avoid issues with
all_species_tree_brooding_pruned$edge.length <- all_species_tree_brooding_pruned$edge.length/all_species_tree_brooding_pruned1$edge.length

#Fit ARD model for brooding behaviour for all species onto the tree:
set.seed(1931)
all_species_brooding_mb_sb_simmap <- make.simmap(tree = all_species_tree_brooding_pruned, 
                                           x = all_species_brooding_mb_sb, 
                                           model = 'ARD', nsim = 10000)

#save simmap object to prevent needing to redo the analysis:
saveRDS(all_species_brooding_mb_sb_simmap, 'Data/all_species_brooding_mb_sb_simmap.rds')

#store all the trees simulated in the simmap creation, this will make it easier to plot the uncertainty at nodes etc.
posterior_all_species_brooding_mb_sb_simmap <- 
  describe.simmap(all_species_brooding_mb_sb_simmap, plot = FALSE)

#Save simmap object as RDS - this is important if the simmap requires output for further analysis
saveRDS(posterior_all_species_brooding_mb_sb_simmap, 'Data/posterior_all_species_brooding_mb_sb_simmap.rds')
```

```{r Plot Simmap object for Binary Brooding Behaviour for All Species}

#import the simmap with all 10000 trees:
posterior_all_species_brooding_mb_sb_simmap <- 
  readRDS('Data/posterior_all_species_brooding_mb_sb_simmap.rds')

pdf("Figures/Mapping Discrete Traits onto Phylogeny/posterior_all_species_brooding_mb_sb_simmap.pdf", width = 8.4, height = 11)

#replace branch lengths back to original (remember we scaled to 1 for model fit)
all_species_tree_brooding_pruned$edge.length <- all_species_tree_brooding_pruned_edges

# Plot reconstruction with posterior probabilities at each node
#colours taken from Wong colour blind palette
brooding_behaviour_colours <- c("mouthbrooder" = "red",
                                "substrate_brooder" = "green"
                                )

 
plot(all_species_tree_brooding_pruned,
     type="fan",
     cex=0.25,
     label.offset = 0.5)
nodelabels(pie=posterior_all_species_brooding_mb_sb_simmap$ace,piecol=brooding_behaviour_colours,
cex=0.25)
# Add the legend
add.simmap.legend(colors=brooding_behaviour_colours, 
                  prompt=FALSE, 
                  fsize=0.75,  
                  size=10, 
                  x = 1,      
                  y = 20)       
add.scale.bar()

dev.off()
```










